Compositional Certified Resource Bounds\ifshort{\vspace{-.15cm}}{}

### 2014-07-07 paper/main.tex
change in abstract
Joshua's remarks for the intro

  This paper describes a novel quantitative resource analysis
  framework that tackles these challenges for C programs.  The
  analysis framework consists of three parts: First, a parametric cost
  semantics formalizes the resource consumption of CompCert Clight
  programs.  Second, a quantitative Hoare logic enables users to
  interactively develop resource bounds in the Coq Proof Assistant.
  The logic is proved sound with respect to the cost semantics, and
  shallow embedding enables a derived bound to be any function that is
  definable in Coq.  Third, an automatic amortized resource analysis
  for C programs computes derivations in the quantitative Hoare logic.
  It is the first automatic amortized analysis that can derive bounds
  that depend on negative integers or differences between numbers,
  which is crucial to handle typical systems code.  Both, the
  quantitative logic and the automatic amortized analysis are
  naturally compositional and can be combined to semi-automatically
  derive global resource bounds.
  %
  An experimental evaluation demonstrates the practicability of the
  analysis framework.  The expressivity of the logic is shown by
  manually deriving customized bounds that are tailored to specific
  algorithms.  A comparison of the automatic amortized analysis with
  other automatic tools on 30 challenging loop and recursion patterns
  from the literature and open-source software shows that the bounds
  derived by the automatic amortized analysis are often more precise.
\end{abstract}

@@

Static analysis of quantitative properties of imperative programs is
an active research area and recent years have seen many innovations.
Notable tools that have been developed include
SPEED~\cite{GulwaniMC09}, KoAT~\cite{BrockschmidtEFFG14},
PUBS~\cite{AlbertAGPZ12}, Rank~\cite{AliasDFG10}, and LOOPUS~\cite{SinnZV14}.
%
While the these tools can derive impressive results for realistic
software, there still exist shortcomings that hamper the application
of existing techniques in practice.
\itemskip
\begin{itemize}
\item Analysis tools are black boxes that \iffull{either}{} deliver a result or fail
  without enabling \emph{user interaction} to manually or
  semi-automatically derive bounds for challenging parts of the
  program.
\itemskipIn
\item The computed bounds are often \emph{non-compositional} local
  bounds (for a single (nested) loop) that are difficult to combine to
  global whole program bounds.
\itemskipIn
\item Existing techniques often rely on complex external tools such as
  abstract interpretation--based invariant
  generation~\cite{GulwaniMC09} or translation of the program into a
  term-rewriting system~\cite{BrockschmidtEFFG14, SinnZV14} without
  providing \emph{verifiable certificates} for the correctness of the
  derived bound.
\end{itemize}
\itemskip
%
While there has been much progress in static quantitative analysis,
Knuth correctly points out in a recent interview~\cite{KnuthInter}
that the state-of-the-art in formal quantitative methods still falls
short in comparison with semantic techniques.
%
\itemskip
\begin{quote}
  % Consider, for example, a computer program that sorts a list of
  % numbers into order. 
  $[\cdots]$ Thanks to the work of Floyd, Hoare, and others,
  we have formal definitions of semantics, and tools by which we can
  verify that sorting is indeed always achieved. My job is to go
  beyond correctness, to an analysis of such things as the program's
  running time $[\cdots]$. I'm 100\% sure that my recurrence correctly
  describes the program's performance, and all of my colleagues agree
  with me that the recurrence is "obviously" valid. Yet I have no
  formal tools by which I can prove that my recurrence is right. I
  don't really understand my reasoning processes at all!
  \vspace{-4ex}
  \begin{flushright}
    -- Donald E. Knuth, 2014
  \end{flushright}
\end{quote}
\itemskip
%
In this work, we develop a resource analysis framework for C programs
that is based on a solid semantic foundation.  The choice of C is
primarily motivated by \iffull{our}{the} ongoing work on the formally verified
hypervisor kernel CertiKOS~\cite{GuVFSC11}.  CertiKOS is mainly
developed in C and is supposed to provide verified guaranties on
timing and memory usage.  Moreover, C is a natural choice because it
is still the most widely used language for system development; in
particular in safety-critical embedded and real-time systems where
resource bound analyses are often required by regulatory
authorities~\cite{May2013}.  Our \emph{contributions} are as follows.
%
\itemskip
\begin{enumerate}
\item We define a operational cost semantics for CompCert Clight that
  defines the resource consumption of terminating and diverging
  executions.  The cost is parametric in a user-definable cost metric
  and can be negative if resources are released during an execution.

\item We develop a quantitative Hoare logic for interactively deriving
  resource bounds for Clight programs.  The logic is implemented and
  proved sound in the Coq Proof Assistant with respect to the cost
  semantics of Clight.
\itemskipIn
\item We describe an automatic static analysis that computes bounds
  together with derivations in the quantitative Hoare logic.  The
  analysis is inspired by type-based amortized resource analysis for
  functional programs~\cite{Jost03,HoffmannAH11}.  It computes
  derivations by generating simple linear constraints that can be
  solved by an off-the-shelf LP solver and does not require any
  fixpoint computations to obtain loop invariants.
\itemskipIn
\item We show with a publically available prototype implementation and
  an experimental evaluation that our novel amortized resource
  analysis works precisely and efficiently for challenging loop and
  recursion patterns, and that the derived constant factors in the
  bounds are close to or identical with the optimal ones.
% \item We prove the completeness of the program logic
%   with respect to the operational cost semantics.
\end{enumerate}
\itemskip
%
To the best of our knowledge, this article presents the first resource
analysis framework for C that makes it possible to combine non-trivial
automatically derived bounds with interactively derived bounds in a
proof system that produces verifiable certificates for the bounds.
Our approach complements existing work since it provides a semantic
foundation for the computation of bounds while still providing support
for automation that sometimes goes beyond the capabilities of existing
techniques.  Both the automatic amortized analysis and the
quantitative logic are naturally compositional and describe the
resource behavior of code fragments without referring to the source
code.  As a result, we directly derive global bounds that are
functions of the input parameters of the program.  Another unique
feature of our system is, that it can handle resources like memory
that may become available during the execution.

@@

Despite the apparent simplicity of the new analysis system, it is able
to reproduce results from the literature (e.g,
SPEED~\cite{GulwaniMC09}) that have been obtained using sophisticated
abstract interpretation--based methods such as disjunctive invariant
generation~\cite{GulwaniMC09} and symbolic backward
execution~\cite{GulwaniZ10}.  In contrast with abstract
interpretation--based methods, our technique does not require any
fixpoint computations to obtain loop invariants.  The mechanism we
designed is able to leverage local assertions such as $x < y$ that we
collect along the branching points of the program to obtain global
resource invariants.  We achieve this by generating a linear
constraint system that reflects resource cost and size changes of
integer variables in the program.  A solution of the linear program
immediately yields a resource usage bound for the C
program.  % We can naturally, handle
% advanced control flow such as break, return, iteration over negative
% integer variables, and mutually-recursive functions.

@@

We have evaluated the automatic analysis with system code and examples
from the literature.  \iffull{\pref{app:cat}}{The extended version
  of this article~\cite{anon_extended}}{} contains more than $30$
challenging loop and recursion patterns that we collected from open
source software and the literature.  Our analysis can find
asymptotically tight bounds for all but 3 of these patterns and in
most cases the derived constant factors are tight.  To compare our
automatic analyzer with existing techniques, we tested our examples
with the tools such as KoAT~\cite{BrockschmidtEFFG14},
Rank~\cite{AliasDFG10}, and LOOPUS~\cite{SinnZV14}.  Our experiments
show that the bounds that we derive are often more precise than those
derived by existing tools.  Only LOOPUS~\cite{SinnZV14}, which also
uses amortization techniques, is able to achieve a similar precision.
%
Several micro benchmarks demonstrate the practicability and
expressiveness of the quantitative Hoare logic.  For example, we
derive a logarithmic bound for a binary search function, a bound that
depends on the contents of an array to describe the exact cost of a
function that finds the maximal element in an array, and a linear
bound that amortizes the cost of $n$ successive increments to a binary
counter.

>>>>

  This paper describes a novel quantitative resource analysis
  framework that tackles these challenges for C programs.  The
  analysis framework consists of three parts: First, a parametric cost
  semantics formalizes the resource consumption of CompCert Clight
  programs.  Second, a quantitative Hoare logic enables users to
  interactively develop resource bounds in the Coq Proof Assistant.
  The logic is proved sound with respect to the cost semantics, and
  shallow embedding enables a derived bound to be any function that is
  definable in Coq.  Third, an automatic amortized resource analysis
  for Clight computes derivations in the quantitative Hoare logic.
  It is the first automatic amortized analysis that can derive bounds
  that depend on negative integers or differences between numbers,
  which is crucial to handle typical systems code.  Both, the
  quantitative logic and the automatic amortized analysis are
  naturally compositional and can be combined to semi-automatically
  derive global resource bounds.
  %
  An experimental evaluation demonstrates the practicability of the
  analysis framework.  The expressivity of the logic is shown by
  manually deriving customized bounds that are tailored to specific
  algorithms.  A comparison of the automatic amortized analysis with
  other automatic tools on 30 challenging loop and recursion patterns
  from the literature and open-source software shows that the bounds
  derived by the automatic amortized analysis are often more precise.
\end{abstract}

@@

Static analysis of quantitative properties of imperative programs is
an active research area and recent years have seen many innovations.
Notable tools that have been developed include
SPEED~\cite{GulwaniMC09}, KoAT~\cite{BrockschmidtEFFG14},
PUBS~\cite{AlbertAGPZ12}, Rank~\cite{AliasDFG10}, and LOOPUS~\cite{SinnZV14}.
%
While the these tools can derive impressive results for realistic
software, there still exist shortcomings that hamper the application
of existing techniques in practice:
\itemskip
\begin{itemize}
\item Analysis tools are black boxes that \iffull{either}{} deliver a result or fail
  without enabling \emph{user interaction} to manually or
  semi-automatically derive bounds for challenging parts of the
  program.
\itemskipIn
\item The computed bounds are often \emph{non-compositional} local
  bounds (for a single (nested) loop) that are difficult to combine to
  global whole program bounds.
\itemskipIn
\item Existing techniques often rely on complex external tools such as
  abstract interpretation--based invariant
  generation~\cite{GulwaniMC09} or translation of the program into a
  term-rewriting system~\cite{BrockschmidtEFFG14, SinnZV14} without
  providing \emph{verifiable certificates} for the correctness of the
  derived bound.
\end{itemize}
\itemskip
%
While there has been much progress in static quantitative analysis,
Knuth correctly points out in a recent interview~\cite{KnuthInter}
that the state-of-the-art in formal quantitative methods still falls
short in comparison with semantic techniques.
%
\itemskip
\begin{quote}
  % Consider, for example, a computer program that sorts a list of
  % numbers into order. 
  $[\cdots]$ Thanks to the work of Floyd, Hoare, and others,
  we have formal definitions of semantics, and tools by which we can
  verify that sorting is indeed always achieved. My job is to go
  beyond correctness, to an analysis of such things as the program's
  running time $[\cdots]$. I'm 100\% sure that my recurrence correctly
  describes the program's performance, and all of my colleagues agree
  with me that the recurrence is "obviously" valid. Yet I have no
  formal tools by which I can prove that my recurrence is right. I
  don't really understand my reasoning processes at all!
  \vspace{-4ex}
  \begin{flushright}
    -- Donald E. Knuth, 2014
  \end{flushright}
\end{quote}
\itemskip
%
In this work, we develop a resource analysis framework for C programs
that is based on a solid semantic foundation.  The choice of C is
primarily motivated by \iffull{our}{the} ongoing work on the formally verified
hypervisor kernel CertiKOS~\cite{GuVFSC11}.  CertiKOS is mainly
developed in C and is supposed to provide verified guaranties on
timing and memory usage.  Moreover, C is a natural choice because it
is still the most widely used language for system development; in
particular in safety-critical embedded and real-time systems where
resource bound analyses are often required by regulatory
authorities~\cite{May2013}.  Our \emph{contributions} are as follows:
%
\itemskip
\begin{enumerate}
\item We define a operational cost semantics for CompCert Clight that
  defines the resource consumption of terminating and diverging
  executions.  The cost is parametric in a user-definable cost metric
  and can be negative if resources are released during an execution.

\item We develop a quantitative Hoare logic for interactively deriving
  resource bounds for Clight programs.  The logic is implemented and
  proved sound in the Coq Proof Assistant with respect to the cost
  semantics of Clight.
\itemskipIn
\item We describe an automatic static analysis that computes bounds
  together with derivations in the quantitative Hoare logic.  The
  analysis is inspired by type-based amortized resource analysis for
  functional programs~\cite{Jost03,HoffmannAH11}.  It computes
  derivations by generating simple linear constraints that can be
  solved by an off-the-shelf LP solver, and does not require any
  fixpoint computations to obtain loop invariants.
\itemskipIn
\item We show with a publically available prototype implementation, and
  an experimental evaluation, that our novel amortized resource
  analysis works precisely and efficiently for challenging loop and
  recursion patterns, and that the derived constant factors in the
  bounds are close to or identical with the optimal ones.
% \item We prove the completeness of the program logic
%   with respect to the operational cost semantics.
\end{enumerate}
\itemskip
%
To the best of our knowledge, this article presents the first resource
analysis framework for C that makes it possible to combine non-trivial
automatically derived bounds with interactively derived bounds in a
proof system that produces verifiable certificates for the bounds.
Our approach complements existing work since it provides a semantic
foundation for the computation of bounds while still providing support
for automation that sometimes goes beyond the capabilities of existing
techniques.  Furthermore, the automatic amortized analysis and the
quantitative logic are naturally compositional and describe the
resource behavior of code fragments without referring to the source
code.  As a result, we directly derive global bounds that are
functions of the input parameters of the program.  Another unique
feature of our system is, that it can handle resources like memory
that may become available during execution.

@@

Despite the apparent simplicity of the new analysis system, it is able
to reproduce results from the literature (e.g,
SPEED~\cite{GulwaniMC09}) that have been obtained using sophisticated
abstract interpretation--based methods such as disjunctive invariant
generation~\cite{GulwaniMC09} and symbolic backward
execution~\cite{GulwaniZ10}.  In contrast with abstract
interpretation--based methods, our technique does not require any
fixpoint computations to obtain loop invariants.  The mechanism we
designed is able to leverage local assertions such as $x < y$ that we
collect along the branching points of the program to obtain global
resource invariants.  We achieve this by generating a linear
constraint system that reflects resource cost and size changes of
integer variables in the program; a solution of the linear program
immediately yields a resource usage bound for the C
program.  % We can naturally, handle
% advanced control flow such as break, return, iteration over negative
% integer variables, and mutually-recursive functions.

@@

We have evaluated the automatic analysis with system code and examples
from the literature.  \iffull{\pref{app:cat}}{The extended version
  of this article~\cite{anon_extended}}{} contains more than $30$
challenging loop and recursion patterns that we collected from open
source software and the literature.  Our analysis can find
asymptotically tight bounds for all but 3 of these patterns, and in
most cases the derived constant factors are tight.  To compare our
automatic analyzer with existing techniques, we tested our examples
with the tools such as KoAT~\cite{BrockschmidtEFFG14},
Rank~\cite{AliasDFG10}, and LOOPUS~\cite{SinnZV14}.  Our experiments
show that the bounds that we derive are often more precise than those
derived by existing tools.  Only LOOPUS~\cite{SinnZV14}, which also
uses amortization techniques, is able to achieve a similar precision.
%
Several micro benchmarks demonstrate the practicability and
expressiveness of the quantitative Hoare logic.  For example, we
derive a logarithmic bound for a binary search function, a bound that
depends on the contents of an array to describe the exact cost of a
function that finds the maximal element in an array, and a linear
bound that amortizes the cost of $n$ successive increments to a binary
counter.
